---
title: "Reproducible Analysis of NOAA Weather Related  Damages to Humans, Propoerty and Agriculture"
output:
  html_document:
    keep_md: yes
  pdf_document: default
---

##The purpose of the assignment is to learn to publish reproducible research using NOAA weather-dsaster data on its impacts on human health, property and agriculture

##Snopisis:NOAA zipped data was read into and unzipped in R, processesed with justifications to get what we need related to the health & economic impacts. In the following steps I will show how I prioriised and subset data with the greatest impacts 
##Steps:
##Processing for Question 1: Steps 1-3
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#1.Process previously downloaded "StormData.csv.bz2"
```{r include=T}
library("R.utils")
library("data.table")
# Here's where we unzip the original file. 
# By setting remove = FALSE you make use of the original file unzipped file 
if(!file.exists("repdata-data-StormData.csv")){
  R.utils::bunzip2("repdata-data-StormData.csv.bz2", remove = FALSE)}
# By setting data.table = FALSE you get a simple data.frame
df<-data.table::fread("repdata-data-StormData.csv", data.table = FALSE)
# Take only what we need to answer both questions
# Jutification: doing this early will speed up the use of CPU memory
df<-df[,c("EVTYPE","FATALITIES","INJURIES","PROPDMG","PROPDMGEXP","CROPDMG","CROPDMGEXP")] 
```

#2.Aggregate and PRIORITISED TOP 15 EVENTS leasing to fatalities & injuries
```{r include=FALSE}
# Take what we need to answer the injuries question
injuries<-df[,c(1:3)]
# aggregate by type of weather event
ecast<-aggregate(.~EVTYPE, injuries, sum)
# oder the injuries and fatalities to take the TOP 15
# justification is we can focus the plotting on the most important damage
inj<-head(ecast[order(ecast$INJURIES, decreasing=T),], n=15)
fat<-head(ecast[order(ecast$FATALITIES, decreasing=T),], n=15)
```

#3.Plot fatalities
# Justifiation to only plot fatalities is that it shows the most serious health effect # On the other hand There's no way of knowing if the injuries a slight or severe
```{r include=T}
library("ggplot2")
qplot(EVTYPE, FATALITIES, data=fat)+ geom_bar(stat="identity")+ ylab("No. of Fatalities") + xlab("Enviromental Disaster Type") + ggtitle("Estimates of Fatalities") + theme(axis.text.x = element_text(angle=80, hjust=1))
```

##RESULTS & DISCUSSION: Question 1: HEALTH EFFECTS
# Re-emphasize the decision to only plot fatalities is that it shows the most serious health effect
# The Top3 categories are : Tornado; Excessive heat & heat in general; and Flash floods & floods in general

##Processing for Question 2: Steps 4-8
#4.Table all the diffeent types of exponents including NAs ("") with some helpful info<<https://rstudio-pubs-static.s3.amazonaws.com/58957_37b6723ee52b455990e149edde45e5b6.html>>
```{r include=T}
table(df$PROPDMGEXP)
table(df$CROPDMGEXP)
```

#5.Sum up the exponets I prioritised to use
````{r include=T}
# Following shows the sum of most significant exponents
# I plan to consider this sub-set as insignificant: c("", "?", "-1", "0")
# Justification is that it save only the processing of 
x <- setdiff(df$PROPDMGEXP, c("", "?", "-1", "0"))
sum(df$PROPDMGEXP %in% x)
y <- setdiff(df$CROPDMGEXP, c("", "?", "-1", "0"))
sum(df$CROPDMGEXP %in% y)
```

#6.Subset dataframe to have at least one exponent each for property or crop 
```{r include=FALSE}
# "yes"", if an observation has at least 1 of the exponents I prioritised above for prop or crop
df$yes<-ifelse(df$PROPDMGEXP %in% x|df$CROPDMGEXP %in% y,1,0)
# this is the new df of interest
lazydf<-df[df$yes==1,]
lazydf<-lazydf[,c("EVTYPE","PROPDMG","PROPDMGEXP","CROPDMG","CROPDMGEXP")]
#let's get rid of the larger  file now to simplify memory use
df<-NULL
lazydf$yes<-NULL 
```

#7.Multiply damage by exponent  to get value$
```{r include=FALSE}
# thesey first have to made to numeric
lazydf$PROPDMGEXP<-as.numeric(chartr("hHKmMB", "223669", lazydf$PROPDMGEXP))
lazydf$CROPDMGEXP<-as.numeric(chartr("kKmMB", "33669", lazydf$CROPDMGEXP))
lazydf$SUMCROPDMG<-lazydf$CROPDMG*10^lazydf$CROPDMGEXP
lazydf$SUMPROPDMG<-lazydf$PROPDMG*10^lazydf$PROPDMGEXP
lazydf<-lazydf[,c("EVTYPE","SUMPROPDMG","SUMCROPDMG")]
```

#8. Re-arrange, aggregate & sort
```{r include=T}
#using melt & dcast as it seems to give me greater values for this one. In Q1 the aggregate function gave the same result
library("reshape2")
fmelt<-melt(lazydf, id=c("EVTYPE"), measure.vars=c("SUMPROPDMG", "SUMCROPDMG"))
fcast<-dcast(fmelt, EVTYPE~variable,sum, na.rm=T)
property<-head(fcast[order(fcast$SUMPROPDMG, decreasing=T),], n=15)
crop<-head(fcast[order(fcast$SUMCROPDMG, decreasing=T),], n=15)
```

#9 Plot property damage 
```{r include=T}
library("ggplot2")
qplot(EVTYPE, SUMPROPDMG, data=property)+ geom_bar(stat="identity")+ ylab("Property Damage Amount $") + xlab("Enviromental Disaster Type") + ggtitle("Estimates of Property Damange") + theme(axis.text.x = element_text(angle=80, hjust=1))
```

#10. Plot crop damage
```{r include=T}
qplot(EVTYPE, SUMCROPDMG, data=crop)+ geom_bar(stat="identity")+ ylab("Property Damage Amount $") + xlab("Enviromental Disaster Type") + ggtitle("Estimates of Crop Damange") + theme(axis.text.x = element_text(angle=80, hjust=1))
```

##RESULTS & DISCUSSION: Question 2: ECONOMIC EFFECTS
# The TOP 3 categories on property effects are from: Floods all types; Hurricanes; Tornadoes 
# The TOP 3 categories on property effects are from: Droughts; Floods all ypes; Storms
